Servis: A dependently typed DSL for web APIs\\ \large Honours Thesis

### 2016-08-25 paper/paper.tex
FINAL FINAL FINAL DRAFT

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\makeatletter
\renewcommand*{\thelisting}{\thesection.\arabic{listing}}
\@addtoreset{listing}{section}
\makeatother
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs}
\begin{document}
\maketitle
\begin{abstract}
In this paper we describe a DSL for web APIs named Servis. Servis is  inspired by Servant, a haskell type-level DSL for web APIs. We identify several issues with the current implementation of Servant and show how Servis solves these by using features of the dependently typed language called Idris.

@@

Though haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values. There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris. This makes doing computations with types a lot more natural. No special constructs are needed like special syntax for type-families; you can just write functions like you would normally do with term-level values. It also allows us to do more powerful things like letting types actually depend on runtime user input.

@@

\subsection{Contributions}
In this paper, I make the following contributions:

@@

\subsection{Explicit status codes}
It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.
It is important for a developer to know how an API call could fail, such that he can handle it sufficently. Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.

This would not be hard to add to Servant. A type-level list of error types paired with the return type of an API endpoint would suffice. However, it would also be interesting to explore if we can enforce that the code that implements a server for an API, actually only returns the error codes that are documented.

\section{Servis}
In this section, I present a DSL in Idris named Servis.
It tries to address each issue listed in Section~\ref{sec:issues}.
Firstly, we will define the DSL in which we can write API specifications. Then, we will look at how to write an interpreter of this DSL that generates documentation.
Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@.
Next, we'll look at solving the overlapping routes problem by using Idris's automatic proof search.
Finally, we will extend the DSL to add dependent types, allowing us to describe dependencies in our API\@.
For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.  It would then force the implementor of the API to only return values smaller or equal to limit in size.

@@

Sadly enough, this code will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
However, this exact problem can be solved with universes We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
the \mintinline{idris}{User} type as a data constructor \mintinline{idris}{USER : ResponseUniverse}. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}, and whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.

@@

\subsubsection{Path and PathPart}
Of course we do not only get request information from the request body,
we also get information about a request by parsing it's request URL. This
is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
When we talk about path parts, we mean sections of a path template that convey parsable information.
For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
in the path \texttt{users}, a variable capture \texttt{user\_id} and a query param \texttt{search}. This
decomposition of a path template can be written down as a datatype. In this datatype we do not only
want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
over a universe similarly to how we defined \mintinline{idris}{Handler}.
\begin{minted}{idris}

@@

\subsection{A simple interpreter: Documentation generation}
Writing an interpreter for generating documentation is straightforward. Our API description isn't some type-level definition, it's just a datatype that can be passed to functions and can be manipulated.

We start by introducing an interface for docs, which just tells us how to turn something into a string:
\begin{minted}{idris}
interface HasDocs a where
  docs :: a -> String
\end{minted}
Now, we just need to define an implementation for \mintinline{idris}{API} for this interface and we are done. To implement this, we will of course also needs to implement the interface for \mintinline{idris}{Handler}, \mintinline{idris}{PathPart} and \mintinline{idris}{Path}.

To get an idea on how this would work, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}:
\begin{minted}{idris}
(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
  docs (GET responseType) =
    "## GET\nResponse type: " ++ docs responseType

@@

We will more-or-less use this representation to model our HTTP handlers.An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.

@@

The router we are going to build uses the same mechanism as described in the Servant paper. In order to be able to write a router, we will have
to modify the definition of \mintinline{idris}{Application} slightly~\cite{servant}. We need to be able to express that the router can not
match the API path with the incoming request. In case of a mismatch, the router should then try if the next available path does match the request.
Instead of letting the application return a value of \mintinline{idris}{IO String}, let it return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just a} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.

@@

The first part of the API for which we will implement the routing mechanics is the \mintinline{idris}{Handler}.
Defining this router consists of two cases: A \mintintline{idris}{GET} or a \mintinline{idris}{POST}.
During a \mintinline{idris}{GET}, we just have to execute the handler which will then give us a response. 
For a \mintinline{idris}{POST}, we need to read the request body from the request, decode it to the right type, and
apply it to the handler to get a response:
\begin{minted}{idris}
(FromRequest req, ToResponse res) => HasServer (Handler req res) where
  -- route : Handler req res -> 
  --         (handler : IO (el res)) ->
  --         URL -> Maybe String -> Maybe (IO String)
  ||| simply turn an (IO (el res)) into an (IO String) 
  ||| using toResponse
  route (GET responseType) handler url requestBody =
    Just (map (toResponse responseType) handler)
  route (POST requestType responseType) handler url requestBody = do
    -- see if there is a requestBody. otherwise fail
    body <- requestBody
    -- try to decode the request. this might fail and return Nothing
    request <- fromRequest requestType body
    -- apply the decoded request to the handler, and encode the result
    Just (map (toResponse responseType) (handler request))
\end{minted}

@@

Routing multiple paths is easy. We just keep on trying consecutive paths until one
does nott return \mintinline{idris}{Nothing}:
\begin{minted}{idris}
( FromCapture capture
, FromQueryParam query
, FromRequest req
, ToResponse resp) =>
  HasServer (API capture query req resp) where

@@

We will now show how we would implement for \mintinline{idris}{GET}:
\begin{minted}{idris}
  (Universe req, Universe res) => Universe (Handler req res) where
    el (GET req res statuses) =
      (DPair Int (\n => Elem n statuses), IO (el responseType))
\end{minted}
So now instead of just returning the IO action, we also return the status code, paired with its proof:
\begin{minted}{idris}
api : Handler EmptyU EmptyU
api = GET VOID [200,401,404]

handler : el api
handler = (MkDPair 401 (the401IsInStatusCodes), putStrLn "hey")
\end{minted}
We can now modify the router such that it can use the status code to write it to the response:
\begin{minted}{idris}
  route (GET responseType statuses)
        (MkDPair status proof, handler)
        url requestBody =
    Just $ do
      putStrLn (show status)
      map (toResponse responseType) handler
\end{minted}
\subsection{A proof of non-overlapping routes}
One of the weaknesess of Servant is that it is possible to define routes that overlap. It would be great if we could give a compiler error in Servis if two routes overlap.

In Idris, we can describe propositions using types. And then allows us to make proofs of these propositions by implementing a program of that type.  Also, Idris comes with a proof searcher that can automatically search for a proof for a certain proposition. It will do a best effort to find a proof, and if it can not find a proof in time the compiler halts with an error. If it did not find a proof, it does not mean there is no possible proof though. The proof searcher is naive. It will only search 100 levels of recursion deep, and it will not do contradictory proofs. So in some cases, the user will have to provide a proof himself.

@@

Now we only have to be concerned with \mintinline{idris}{Const} and \mintinline{idris}{Capture}. Because \mintinline{idris}{Capture} signifies a variable substitution, it will always conflict with a \mintinline{idris}{Const}. E.g.,  \texttt{/users/hey} conflicts with \texttt{/users/:var} because \texttt{:var} could be substituted by \texttt{hey}.

This leads us to defining the following datatype, which allows us to prove that a \mintinline{idris}{PathPart} is disjoint if  two \mintinline{idris}{Const} values have different names:
\begin{minted}{idris}
data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
  ConstD : IsNo (decEq str str') -> DisjointPP (Const str) (Const str')
\end{minted}

@@

The fact that we can not construct a value of \mintinline{idris}{DisjointPP (Const x) (Capture x y)} indicates that a \mintinline{idris}{Const}, and a \mintinline{idris}{Capture} always overlap, which is the behaviour that we wanted to express.

@@

However, if we introduce an overlap, the compiler will tell us that this
api definition is invalid and that we need to fix our code:
\begin{minted}{idris}
apiOverlap : List (Path CapU EmptyU EmptyU RespU)
apiOverlap
  =  Const "users" :> Capture "name" STRING :> Outputs (GET USER)
  +> Const "users" :> Const "all"           :> Outputs (GET (LIST USER))
  +> Nil

@@

@@

\subsubsection{New problems that dependent types create}
The dependent types are nice, but our datatype is now a Higher-order abstract syntax instead of a first-order abstract syntax.
Which means the datatype contains a function.  This makes it difficult to generate documentation as we need to provide a value
to apply to the function to traverse the datatype. It isn't impossible though. We could introduce the following interface:
\begin{minted}{idris}
interface Universe u => HasSample u where
  sample : (v : u) -> el v

@@

The real problem though is that this representation breaks the overlapping path detector.
The overlapping path detector uses the feature of idris to search proofs automatically. But
proof search isn't perfect will be exhausted after a while. Take ourexample with the
\mintinline{idris}{limit : Nat} parameter. For Idris to find a proof, it would try every number
in \mintinline{idris}{Nat} to  make sure that the part of our path after the dependency does
not overlap. However, there are a countable infinite number of \mintinline{idris}{Nat}s so
Idris would never be able to complete the proof. Hence the auto-searcher will stop searching
for a proof after a while and quit.

@@

We were able to solve several shortcomings of servant: By not using an open kind, 
we eliminated the possibility of writing APIs that are not interpretable.  We also also the possibility
of defining overlapping routes. W made status codes explicit in the API and gave
library authors the tools to prove that their servers adhere to these status code
definitions.  Finally, we were able to implement an dependent API, that allows interdependencies
within paths to be able to describe even more powerful APIs.

\section{Related work}
This project started with me looking at\url{https://github.com/tel/serv}. Which is
a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
what would happen if we would write something similar Idris.  The serv source code is kind of
dense because Haskell isn't a very nice language to do type-level programming in
and it got me curious whether or not Idris would improve readability. Which it did!

Also, once I had a working example of Serv in idris, I wanted to see if idris
could add any new tricks to it. This made me implement dependent types and
route conflict resolution.
\bibliography{sources}
\bibliographystyle{plain}
\end{document}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\makeatletter
\renewcommand*{\thelisting}{\thesection.\arabic{listing}}
\@addtoreset{listing}{section}
\makeatother
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs\\
\large Honours Thesis}
\begin{document}
\maketitle
\begin{abstract}
In this paper we describe a DSL for web APIs named Servis. Servis is  inspired by Servant, a Haskell type-level DSL for web APIs. We identify several issues with the current implementation of Servant and show how Servis solves these by using features of the dependently typed language called Idris.

@@

Though Haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values. There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris. This makes doing computations with types a lot more natural. No special constructs are needed like special syntax for type-families; you can just write functions like you would normally do with term-level values. It also allows us to do more powerful things like letting types actually depend on runtime user input.

@@

\subsection{Contributions}
In this paper, we make the following contributions:

@@

\subsection{Explicit status codes}
It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.

It is important for a developer to know how an API call could fail, such that he can handle it accordingly. Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.

This would not be hard to add to Servant. A type-level list of error types paired with the return type of an API endpoint would suffice. However, it would also be interesting to explore if we can enforce that the code that implements a server for an API actually only returns the error codes that are documented.

\section{Servis}
In this section, I present a DSL in Idris named Servis.
It tries to address each issue listed in Section~\ref{sec:issues}.
Firstly, we will define the DSL in which we can write API specifications in a closed kind so that no malformed
descriptions can exist. Then, we will look at how to write an interpreter of this DSL that generates documentation.
Next, we will look at solving the overlapping routes problem by using Idris's automatic proof search.
Then, we will  extend the DSL to define accepted status coded, and explore a way to check if handlers adhere to this definition.
Finally, we will extend the DSL to add dependent types, allowing us to describe dependencies in our API\@.

@@

Sadly enough, this code will give us a compiler error. The problem is that in Idris, we cannot pattern match on values of \mintinline{idris}{Type}. 
However, this exact problem can be solved with universes We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
the \mintinline{idris}{User} type as a data constructor \mintinline{idris}{USER : ResponseUniverse}. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}, and whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.

@@

\subsubsection{Path and PathPart}
Of course we do not only get request information from the request body,
we also get information about a request by parsing its request URL. This
is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
When we talk about path parts, we mean sections of a path template that convey parsable information.
For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
in the path \texttt{users}, a variable capture \texttt{user\_id} and a query param \texttt{search}. This
decomposition of a path template can be written down as a datatype. In this datatype we do not only
want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
over a universe similarly to how we did in the definition of \mintinline{idris}{Handler}.
\begin{minted}{idris}

@@

\subsection{A simple interpreter: Documentation generation}
Writing an interpreter for generating documentation is straightforward. Our API description is not just some type-level definition, it is  a datatype that can be passed to functions and can be manipulated.

We start by introducing an interface for docs, which simply tells us how to turn something into a string:
\begin{minted}{idris}
interface HasDocs a where
  docs :: a -> String
\end{minted}
Now, we need to define an implementation for \mintinline{idris}{API} for this interface and we are done. To implement this, we will of course also needs to implement the interface for \mintinline{idris}{Handler}, \mintinline{idris}{PathPart} and \mintinline{idris}{Path}.

To get an idea on how this would loook like, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}. The others are left as an exercise for the reader:
\begin{minted}{idris}
(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
  docs (GET responseType) =
    "## GET\nResponse type: " ++ docs responseType

@@

We will more-or-less use this representation to model our HTTP handlers. An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskell's \texttt{warp} web server (which impelements this concept) to Idris.

@@

The router we are going to build uses the same mechanism as described in the Servant paper. In order to be able to write a router, we will have
to modify the definition of \mintinline{idris}{Application} slightly~\cite{servant}. We need to be able to express that the router can not
match the API path with the incoming request. In case of a mismatch, the router should then try if the next available path does match the request.
Instead of letting the application return a value of \mintinline{idris}{IO String}, we let it return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just a} signals a route match, and \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.

@@

The first part of the API for which we will implement the routing mechanics is the \mintinline{idris}{Handler}.
Defining this router consists of two cases: A \mintinline{idris}{GET} or a \mintinline{idris}{POST}.
During a \mintinline{idris}{GET}, we just have to execute the handler which will then give us a response. 
For a \mintinline{idris}{POST}, we need to read the request body from the request, decode it to the right type, and
apply it to the handler to get a response:
\begin{minted}{idris}
(FromRequest req, ToResponse res) => HasServer (Handler req res) where
  -- route : Handler req res -> 
  --         (handler : IO (el res)) ->
  --         URL -> Maybe String -> Maybe (IO String)
  ||| simply turn an (IO (el res)) into an (IO String) 
  ||| using toResponse
  route (GET responseType) handler url requestBody =
    Just (map (toResponse responseType) handler)
  route (POST requestType responseType) handler url requestBody = do
    -- see if there is a requestBody. otherwise fail
    body <- requestBody
    -- try to decode the request. this might fail and return Nothing
    request <- fromRequest requestType body
    -- apply the decoded request to the handler, and encode the result
    Just (map (toResponse responseType) (handler request))
\end{minted}

@@

Routing multiple paths is easy. We just keep on trying consecutive paths until one
does not return \mintinline{idris}{Nothing}:
\begin{minted}{idris}
( FromCapture capture
, FromQueryParam query
, FromRequest req
, ToResponse resp) =>
  HasServer (API capture query req resp) where

@@

\subsection{A proof of non-overlapping routes}
One of the weaknesess of Servant is that it is possible to define routes that overlap.
It would be great if we could give a compiler error in Servis if this is the case.

In Idris, we can describe propositions using types. This then allows us to make proofs of these propositions by implementing a program of that type.  Idris also comes with a proof searcher that can automatically search for a proof for a certain proposition. It will do a best effort to find a proof, and if it can not find a proof in time the compiler halts with an error. If it did not find a proof, it does not mean there is no possible proof though. The proof searcher is naive. It will only search 100 levels of recursion deep, and it will not do contradictory proofs. So in some cases, the user will have to provide a proof himself.

@@

Now we only have to be concerned with \mintinline{idris}{Const} and \mintinline{idris}{Capture}. Because \mintinline{idris}{Capture} signifies a variable substitution, it will always conflict with a \mintinline{idris}{Const}. E.g.,  \texttt{/users/hey} conflicts with \texttt{/users/:var} because \texttt{:var} could be substituted by \texttt{hey}. It will also conflict with another \mintinline{idris}{Capture}.

This leads us to defining the following datatype, which allows us to prove that a \mintinline{idris}{PathPart} is disjoint if  two \mintinline{idris}{Const} values have different names:
\begin{minted}{idris}
data DisjointPP : PathPart capture query -> PathPart capture query -> Type where
  ConstD : IsNo (decEq str str') -> DisjointPP (Const str) (Const str')

@@

The fact that we can not construct a value of \mintinline{idris}{DisjointPP (Const x) (Capture x y)} nor a \mintinline{idris}{DisjointPP (Capture x y) (Capture z f)} indicates that these will always overlap, which is the behaviour that we wanted to express.

@@

However, if we introduce an overlap, the compiler will tell us that this
API definition is invalid and that we need to fix our code:
\begin{minted}{idris}
apiOverlap : List (Path CapU EmptyU EmptyU RespU)
apiOverlap
  =  Const "users" :> Capture "name" STRING :> Outputs (GET USER)
  +> Const "users" :> Const "all"           :> Outputs (GET (LIST USER))
  +> Nil

@@

We will now show how we would implement this for \mintinline{idris}{GET}:
\begin{minted}{idris}
  (Universe req, Universe res) => Universe (Handler req res) where
    el (GET req res statuses) =
      (DPair Int (\n => Elem n statuses), IO (el responseType))
\end{minted}
So now instead of just returning the IO action, we also return the status code, paired with its proof:
\begin{minted}{idris}
api : Handler EmptyU EmptyU
api = GET VOID [200,401,404]

handler : el api
handler = (MkDPair 401 (the401IsInStatusCodes), putStrLn "hey")
\end{minted}
We can now modify the router such that it can use the status code to write it to the response:
\begin{minted}{idris}
  route (GET responseType statuses)
        (MkDPair status proof, handler)
        url requestBody =
    Just $ do
      putStrLn (show status)
      map (toResponse responseType) handler
\end{minted}

@@

\subsubsection{New problems that dependent types create}
The dependent types are nice, but now our datatype contains a function.
This makes it difficult to generate documentation as we need to provide a value
to apply to the function to traverse the datatype. It is not impossible though. We could introduce the following interface:
\begin{minted}{idris}
interface Universe u => HasSample u where
  sample : (v : u) -> el v

@@

The real problem though is that this representation breaks the overlapping path detector.
The overlapping path detector uses the feature of idris to search proofs automatically. But
proof search is not perfect and will be exhausted after a while. Take our example with the
\mintinline{idris}{limit : Nat} parameter. For Idris to find a proof, it would try every number
in \mintinline{idris}{Nat} to  make sure that the part of our path after the dependency does
not overlap. However, there are a countable infinite number of \mintinline{idris}{Nat}s so
Idris would never be able to complete the proof. Hence the auto-searcher will stop searching
for a proof after a while and quit.

@@

We were able to solve several shortcomings of servant: By not using an open kind, 
we eliminated the possibility of writing APIs that are not interpretable.  We also also the possibility
of defining overlapping routes. W made status codes explicit in the API and gave
library authors the tools to prove that their servers adhere to these status code
definitions.  Finally, we were able to implement an dependent API, that allows interdependencies
within paths to be able to describe even more powerful APIs. However, these dependent types do make
it more difficult to write implementations for APIs.  A users will have to provide proofs that his
handlers adhere to the pre-defined status codes, and will also have to prove manually that his
routes do not overlap.

\section{Related work and Future work}
This project started with me looking at \url{https://github.com/tel/serv}. Which is
a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
what would happen if we would write something similar in a dependently typed language.
I had never used a dependently typed language before, or really thought of what it would gain me.
I decided to learn Idris, and discovered that it would make implementing a Servant alternative
very elegant.

When I got an implementation working, I got curious if I could add more exotic features using Idris.
The question now is whether the things I have discovered could be ported back to serv or Servant 
once dependent types get added to GHC. This would be great to explore in the future.
\bibliography{sources}
\bibliographystyle{plain}
\end{document}
